Likelihoods for a general class of ARGs under the SMC

### 2025-02-21 paper.tex
Another pass

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to 
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, with the uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the 
details of recombination events, and heuristic methods (in different ways)
omit these details. In many ways, the significant progress made in
scalability by these heuristic methods is \emph{because} they don't
attempt to precisely reconstruct recombination events
(this should be taken into account when interpreting the resulting ARGs;
see \citet{wong_general_2023} for more discussion).

@@

A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. 
Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. 
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, the high-level of detail about the underlying
events that is required is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical ms
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The msprime
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to msprime~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} also do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

In this paper we provide a new formulation of ARG likelihood which 
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs.
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we give a backwards-in-time description of the SMC
which naturally leads to a straightforward method for computing the likelihood
of any ARG, and describe an efficient algorithm to do so.
We show that explicitly identifying recombination events via nodes 
in the graph is not necessary as long as the ``locally
unary''~\cite{wong_general_2023,fritze2024forest} sections of common ancestor
nodes are retained, from the perspective of computing the likelihood.
We also show via a empirical examples that the likelihood is remarkably
robust to the presence of polytomies, and illustrate how it can be
decomposed into separately evaluated time slices.
Finally, we discuss how these results may be applied and extended.

@@

\section*{Discussion}
The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial 
and imprecise information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of 
ARG~\citep{wong_general_2023}, 
a significant increase in flexibility over the 
strict input requirements for the classical Kuhner-Yamato-Felsenstein 
formulation. 
We have shown via some illustrative examples that the formulation is 
robust to incomplete and imprecise information, tolerating the loss 
of a significant fraction of the internal nodes in the ARG to minimal effect.
We have also shown that the likelihood function performs well 
(at least in our toy example) on ARGs inferred by tsinfer+tsdate,
successfully capturing information about latent recombination events
from imprecise and noisy data.
Locally unary nodes play a key role, and although not
currently a common focus of ARG research~\citep{wong_general_2023}, 
they are estimated by tsinfer and can be imputed to high a
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
% I think we have to grasp this nettle or we're inviting the reviewers to 
% give us a bunch of "what if you tried this" bits of extra homework.
Fully characterising the properties of the likelihood on the output 
of tsinfer and other methods is beyond the scope of this work.
Similarly, while there are many different potential applications to concrete 
inferential problems, the details are nontrivial and must be
consigned to future work.

@@

Our results, while preliminary, suggest that ``full ARGs'', in which 
each recombination and common ancestor event is precisely estimated,
contain a significant degree of redundancy in terms of capturing 
information about the generative process. 
It is possible, therefore, that explicitly estimating 
\emph{all} such events is not necessary to sample ARGs under the 
SMC (as ARGweaver and SINGER do). 
A combined approach, in which the uncertainty around the
ordering of events that are 
weakly informed by the data is encoded structurally in the 
graph and other forms of uncertainty captured by sampling from 
the posterior, may yield further improvements in 
scalability for statistically rigorous inference.
% Signicant work would be needed, however, to make 
% downstream analysis robust to such stuctural uncertainty;
% polytomies, for example, are often not
Another exciting prospect for ARG inference is the development of 
``hybrid'' inference methods, where hueristic methods such as 
tsinfer or ARG-Needle are used to infer
the details of the recent past in huge datasets such as 
UK Biobank, passing over to a rigorous model-based
approach for the ancient past.
This is similar in spirit to the widely-used approach 
of simulating complex dynamics of the recent past using 
detailed but slow models and using faster coalescent simulations
in the distant past when the assumptions required 
are more justified~\citep{bhaskar_distortion_2014,
kelleher_efficient_2018,haller_tree_2018,nelson_accounting_2020}.
% Maybe we shouldn't bother saying this, but I can't resist the 
% opportunity to say "hey, you should be using tskit!"
% It feels like some PIs aren't getting this message, so need to have
% it spelled out.
Just as forward and backward simulations are useful in different contexts
and can be combined to great effect,
large-scale hueristic and MCMC sampling inference methods have their 
applications, and there may be synergy in their combination. 
% This is pretty weak, tbh.
The practical details of ``handing over'' between methods
will require cooperation and precise communication;
the tskit library, with its well-defined data model and flexible
metadata support, is the ideal basis for this.

>>>>

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to 
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, with the uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the 
details of recombination events, and heuristic methods (in different ways)
omit these details. In many ways, the significant progress made in
scalability by these heuristic methods is \emph{because} they don't
attempt to precisely reconstruct recombination events.

@@

A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. 
Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. 
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, the high level of detail about the underlying
events that is required is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical ms
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The msprime
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to msprime~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} also do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

In this paper we provide a new formulation of ARG likelihood which 
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs.
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we give a backwards-in-time description of the SMC
which naturally leads to a straightforward method for computing the likelihood
of any ARG, and describe an efficient algorithm to do so.
We show that explicitly identifying recombination events via nodes 
in the graph is not necessary 
(from the perspective of computing the likelihood)
as long as the ``locally
unary''~\citep{wong_general_2023,fritze2024forest} sections of common ancestor
nodes are retained.
We also show via an empirical example that the likelihood is 
robust to the presence of polytomies and other imprecise 
information, and illustrate how it can be
decomposed into separately evaluated time slices.
Finally, we discuss how these results may be applied and extended.

@@

\section*{Discussion}
The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial 
and imprecise information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of 
ARG~\citep{wong_general_2023}, 
a significant increase in flexibility over the 
strict input requirements for the classical Kuhner-Yamato-Felsenstein 
formulation. 
Locally unary nodes play a key role, and although not
currently a common focus of ARG research~\citep{wong_general_2023}, 
they are estimated by tsinfer and can be imputed to high a
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
We have shown via some illustrative examples that the formulation is 
robust to incomplete and imprecise information, tolerating the loss 
of a significant fraction of the internal nodes in the ARG to minimal effect.
We have also shown that the likelihood function performs well 
(at least in our toy example) on ARGs inferred by tsinfer+tsdate,
successfully capturing information about latent recombination events
from imprecise and noisy data.
% I think we have to grasp this nettle or we're inviting the reviewers to 
% give us a bunch of "what if you tried this" bits of extra homework.
Theoretical explanations of this apparent robustness,
and fully characterising the properties of the likelihood on the output 
of tsinfer and other methods is an important avenue for future work.
Similarly, while there are many potential applications to concrete 
inferential problems, the details are nontrivial and must be
consigned to future work.

@@

Our results, while preliminary, suggest that ``full ARGs'', in which 
each recombination and common ancestor event is precisely estimated,
contain a significant degree of redundancy in terms of capturing 
information about the generative process. 
It is possible, therefore, that explicitly estimating 
\emph{all} such events is not necessary to sample ARGs under the 
SMC (as ARGweaver and SINGER do). 
A combined approach, in which the uncertainty around the
ordering of events that are 
weakly informed by the data is encoded structurally in the 
graph and other forms of uncertainty captured by sampling from 
the posterior, may yield further improvements in 
scalability for statistically rigorous inference.
% Signicant work would be needed, however, to make 
% downstream analysis robust to such stuctural uncertainty;
% polytomies, for example, are often not
Another exciting prospect for ARG inference is the development of 
``hybrid'' inference methods, where hueristic methods such as 
tsinfer or ARG-Needle are used to infer
the details of the recent past in huge datasets such as 
UK Biobank, passing over to a rigorous model-based
approach for the ancient past.
This is similar in spirit to the widely-used approach 
of simulating complex dynamics of the recent past using 
detailed but slow models and using faster coalescent simulations
in the distant past when the assumptions required 
are more justified~\citep{bhaskar_distortion_2014,
kelleher_efficient_2018,haller_tree_2018,nelson_accounting_2020}.
% Maybe we shouldn't bother saying this, but I can't resist the 
% opportunity to say "hey, you should be using tskit!"
% It feels like some PIs aren't getting this message, so need to have
% it spelled out.
Just as forward and backward simulations are useful in different contexts
and can be combined to great effect,
large-scale hueristic and Bayesian sampling inference methods have their 
applications, and there may be synergy in their combination. 
% This is pretty weak, tbh.
The practical details of ``handing over'' between methods
will require cooperation and precise communication;
the tskit library, with its well-defined data model and flexible
metadata support, is the ideal basis for this.
